## Introduction to dplyr

dplyr is an R package in the tidyverse. We can load the package using

```{r}
install.packages("dplyr")

# load required ;ibraries
library("dplyr")
library(ggplot2)
?dplyr

# preview dataset
head(starwars)
```


### Filtering data
We can filter data using `filter()`. This allows us to subset observations (rows) based on their values (in columns).

Tips:

  * Be sure you spell the column name correctly (an the value name if it's a categorical variable). Remember, R is case-sensitive
  * Be sure to use `==` when comparing observations. (Remember, `=` is an assignment operator)
  * You can use <,>, >=, <= to compare numeric or categorical variables (nominal variables are ranked alphabetically, while ordinal variables have a built-in rank)

```{r}
# or is |
filter(starwars, hair_color == "blond" | eye_color == "blue")
filter(starwars, hair_color %in% c("blond", "blonde")) 

importantPlanets <- c("Tatooine", "Alderaan", "")
filter(starwars, homeworld %in% importantPlanets)

filter(starwars, hair_color == "blond" & species == "Human")
filter(starwars, species == "Human" , homeworld == "Tatooine" , skin_color == "fair")

filter(starwars, mass >= 75, mass <= 100, hair_color == "brown", height > 170)

filter(starwars, mass != 75, name < "Mace")

filterData <- filter(starwars, species == "Human" , homeworld == "Tatooine" , skin_color == "fair")

# Ranked data
head(diamonds) # displays first 6 rows
class(diamonds$cut) # gives you the specific type of data
summary(diamonds$cut) # summary gives you a count of each category or summary statistics if numeric
summary(diamonds$carat)
head(diamonds$cut) # displays first 6 values (and levels)

filter(diamonds, cut > "Good")


## Ordering categorical data
unique(starwars$eye_color)
factor(starwars$eye_color, 
       c("red","orange","gold","yellow","green,yellow", "blue", "black"),
       ordered = T)

## Find all characters that are shorter than 100 cm
filter(starwars, height < 100, species != "Droid") # use != for not equal to or everything excpet

## Which characters were born between the years 100 and 200 (inclusive)
filter(starwars, birth_year >= 100, birth_year <= 200)
filter(starwars, between(birth_year, 100, 200))

## Which characters weigh over 100kg but are shorter than 185 cm
filter(starwars, mass > 100, height < 185)

```

### Using NA
```{r}
NA > 5 #it doesn't know
NA + 10
sum(10,NA,5, na.rm = T)
NA/2



```

### Arranging data
`arrange()` reorders rows. It does not remove any rows.NA values are always at the end when you order by a column

```{r}

# lowest to highest birth_year
arrange(starwars, birth_year)

# highest to lowest birth year
arrange(starwars, desc(birth_year))

# categorical is alphabetical
arrange(starwars,hair_color)

# multiple columns
arrange(starwars,hair_color, birth_year)


### Practice!

## Arrange starwars characters to find the tallest characters and the shortest characters
arrange(starwars, height)
arrange(starwars, desc(height))

## Alhabetize the star wars characters by name
arrange(starwars, name)

## How could you use arrange() sort all missing values to the start?
arrange(starwars, desc(is.na(height)), desc(is.na(mass)))

arrange(starwars, desc(is.na(starwars)))

```

We can select certain columns in the dataset

`select()` allows us to retain only certain variables (columns). It doesn't change the order, but it removes columns not named

```{r}

select(starwars, hair_color, skin_color, eye_color)

head(starwars)
select(starwars, hair_color:eye_color) # resturns every coklumn between first:last

select(starwars, -hair_color)
select(starwars, -(hair_color:eye_color))

starwars_no_color <- select(starwars, -(hair_color:eye_color))
ggplot(starwars_no_color, aes(x = hair_color)) # error because we removed color

selelct(starwars, contains("color"))
select(starwars, ends_with("color"))
select(starwars, contains("_"))
select(starwars, starts_with("s"), ends_with("color"))

?select
?is.na

starwars2 <- rename(starwars, birthYear = birth_year)
starwars2

### Practice

## Select out the homeworld and species for the starwars dataset. What question might this subset of varaibles help answer?
select(starwars, homeworld, species)

## Brainstorm as many ways as possible to select hair_color, eye_color, skin_color, and birth_year
select(starwars, hair_color:birth_year)

select(starwars, ends_with("color"), ends_with("year"))

```

### Adding new column
`mutate()` adds new columns to the end of your dataset.

```{r}

starwars_small <- select(starwars, height, mass, birth_year)
head(starwars_small)

starwars_small <- mutate(starwars_small, height_m = height/100)
mutate(starwars_small, bmi = mass/(height_m^2))

starwars_small <- select(starwars, height, mass, birth_year)
mutate(starwars_small, height_m = height/100, bmi = mass/(height_m^2))

# to only keep new columns, use transmutate
transmute(starwars_small,
          height_m = height/100,
          bmi = mass/(height_m^2))

## using aggregate functions
prop_mass <- mutate(starwars_small, proportional_mass = mass/sum(mass, na.rm = T))
arrange(prop_mass, desc(proportional_mass))

```


### Summarizing and grouping data

`summarize()`collapses an entire column of data to a single value

```{r}

mutate(starwars, mean_mass = mean(mass, na.rm = T))
summarize(starwars, 
          mean_mass = mean(mass, na.rm = T),
          mean_height = mean(height, na.rm = T))

mean(starwars$mass, na.rm = T)

species_masses <- summarize(group_by(starwars,species),
          mean_mass = mean(mass, na.rm = T),
          count = n()) # n() returns the number of values in each species

# the pipe operator %>%
# function(x, y) is the same as x %>% function(y)
# When using dplyr functions, generally always start with dataset
species_masses <- starwars %>%
  group_by(species) %>%
  summarize(mean_mass = mean(mass, na.rm = T),
            count = n()) %>%
  arrange(desc(mean_mass))

arrange(species_masses, desc(mean_mass))

```

### Sampling a designated number of rows
`sample()` allows us to sample a random number of rows from our dataset. Sampling is random

```{r}

# 10 random rows
starwars_10rows <- sample_n(starwars,10)

starwars_10rows

# 10% of rows
nrow(starwars) # total number of rows in 87
starwars_10percent <- sample_frac(starwars, 0.1)
starwars_10percent #9 rows is 10%

starwars %>%
  group_by(gender) %>%
  sample_n(4)

starwars %>%
  group_by(species) %>%
  sample_n(1)

```

We can also take a slice of our dataset using `slice()`and it's related set of functions. This retains the order of our rows in the dataset and is not random.

```{r}

slice(starwars, 50:60) #slice by index value (row number)

slice_head(starwars, n=10) #first 10 rows by index
slice_tail(starwars, n=10) #last 10 rows by index

slice_min(starwars, order_by = mass, n=10) #rows with lowest 10 masses
slice_max(starwars, order_by = mass, n=10) # rows with highest 10 masses



```